\textbf{Socioeconomic deprivation and urban mobility in Latin America after COVID-19}

### 2024-01-30 manuscript/manuscript.tex
manuscript-refs

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\usepackage{lipsum} \usepackage{libertine}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Urban mobility in Latin America after COVID-19},
  pdfauthor={Carmen Cabrera-Arnau; Francisco Rowe; Miguel González-Leonardo; Andrea Nasuto; Ruth Neville},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{\textbf{Urban mobility in Latin America after COVID-19}}

@@

\date{}
\begin{document}
\maketitle
\begin{abstract}
The movement of people between locations is crucial for sustainable and
inclusive cities, facilitating the exchange of knowledge and resources.
Due to restrictions, lockdowns and fear of crowded places, the COVID-19
pandemic significantly reduced the number and extent of people's
movements. Existing research suggests that the impact of the pandemic on
mobility was unequal across locations and socioeconomic groups. However,
evidence and analysis of long-term changes in mobility is scarce,
especially in less developed countries. In this study we use location
data from Meta-Facebook users to analyse patterns of urban mobility in X
Functional Urban Areas from South American countries. Findings reveal a
general decrease in mobility during the pandemic, with gradual recovery
trends up to April 2023. However, socioeconomic disparities persist
through the period of analysis, with deprived areas experiencing smaller
initial losses in mobility. Inequalities in mobility levels across
socioeconomic groups generally diminish over time, although they persist
in all the countries included in the analysis. This research highlights
the importance of understanding the long-term effects of the pandemic on
mobility and the need for equitable policy responses to address societal
challenges in urban areas.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, boxrule=0pt, sharp corners, frame hidden, interior hidden]}{\end{tcolorbox}}\fi

\hypertarget{sec-intro}{%
\section{Introduction}\label{sec-intro}}

@@

A key barrier to monitor changes in geographic mobility patterns in less
developed countries during and post the COVID-19 pandemic has been the
lack of suitable data. Traditionally census and survey data have been
employed to analyse human mobility patterns in these countries. Yet,
these data streams are not frequently updated and suffer from slow
releases, with census data for example being collected over intervals of
ten years in most countries. Traditional data streams thus lack the
temporal granularity to analyse population movements over short-time
periods and offer an up-to-date representation of the urban mobility
system. Data resulting from social interactions on digital platforms
have emerged as an unique source of information to deliver this
representation and capture human population movement in less developed
countries at scale. Particularly location data from mobile phone
applications have become a prominent source to sense patterns of human
mobility at higher geographical and temporal resolution in real time.

Drawing on a dataset of 213 million observations from Meta-Facebook
users' mobile location data, we aim to assess socioeconomic differences
in the extent and persistence of decline in urban mobility in Argentina,
Chile, Colombia and Mexico during and after the COVID-19 pandemic from
March 2020 to March 2023. We use Meta-Facebook observed data to measure
origin-destination flows from March 2020 to May 2022, and Meta Prophet
time-series forecasting machine learning algorithm to predict
origin-destination flows from June 2022 to March 2023. We use Functional
Urban Areas (FUAs) boundaries from the Global Human Settlement Layer
(GHSL) to define urban areas, and the Global Gridded Relative
Deprivation Index (GRDI) developed by NASA's Socioeconomic Data and
Applications Centre (SEDAC) from sociodemographic and satellite data
inputs. Building on existing evidence, we hypothesised that (1) urban
mobility has recovered returning to pre-pandemic levels of movement as
nonpharmaceuthical restrictions were lifted; and (2) socioeconomic
differences in urban mobility have endured the pandemic reflecting deep
societal inequalities as knowledge-intensive businesses adopt hybrid
working.

\begin{itemize}
\tightlist
\item
  Context and importance of human mobility: public health , climate
  change\\
\item
  The pandemic resulted in major changes in human mobility - unequal
  impacts across socioeconomic groups\\
\item
  Gap 1: Little work on understanding the long-term changes in mobility
  - recovery\\
\item
  Gap 2: little work on less developed countries - where inequalities
  are more pronounced\\
\item
  Aim - focus on South American countries as case study.
\end{itemize}

\hypertarget{sec-results}{%
\section{Results}\label{sec-results}}

The evolution of the levels of movement is measured with respect to a
baseline period prior to the pandemic as described in
Section~\ref{sec-methods}. For the purposes of the analysis, we
aggregate the raw movement data temporally into months and spatially
into administrative units according to various GADM levels. The analysis
focuses on administrative areas that are within the boundaries of
functional urban areas as specified by the Global Human Settlement
Layer. For each administrative area, we compute the Relative Deprivation
Index (RDI) based on data from NASA's Socioeconomic Data and
Applications Centre (SEDAC). Figure 1 displays the administrative areas
included in the study, coloured according to the Relative Index of
Deprivation. Predictions about the evolution of the levels of movement
are made using the Prophet forecasting procedure. Further details are
provided in Section~\ref{sec-methods}.

\begin{figure}

@@

\end{figure}

\hypertarget{the-impact-of-covid-19-on-urban-mobility}{%
\subsection{The impact of COVID-19 on urban
mobility}\label{the-impact-of-covid-19-on-urban-mobility}}

We analyse the evolution of the relative intensity of urban mobility
with respect to a baseline period prior to the pandemic. Specifically,
movements covering a distance of at most 70 km are considered. For a
movement to be classified as urban it needs to start or end within a
functional urban area from Argentina, Chile, Colombia and Mexico. The
observed data is available for a two-year period starting in April 2020,
just after the first wave of COVID-19 pandemic cases, and ending in
March 2022. After 2022 no observations are available, however, we
generate a 12-month forecast up to March 2023 in order to gain a better
understanding of the recovery trends.

Figure 2 displays the patterns of recovery for the mobility levels in
the administrative units belonging to functional urban areas in the
countries of interest. The three lines in each panel represent the mean
levels of mobility for administrative units grouped into one of three
terciles, according to their average relative deprivation index.

\begin{figure}

@@

\end{figure}

Generally, there was a drop in the levels of mobility with respect to
the baseline period in all four countries. This drop was especially
large for Argentina, Chile and Colombia, with Mexico displaying a
smaller decrease in the relative number of movements with respect to the
baseline. Following the initial drop in movement, all four countries
evolve towards the recovery of baseline levels of urban mobility, with a
generally increasing trend. There are, however, fluctuations from the
general trend which manifest differently for each country. These
fluctuations mirror each other in the case of Argentina and Colombia,
where urban mobility sharply bounces back closer to pre-pandemic levels
around July of 2020. Chile and Mexico display more progressive patterns
of recovery, although Chile never reaches baseline levels. These
fluctuations are unique to each country and can be attributed to local
factors such as the effects of seasonality or the different stringency
measures imposed by the national governments during the pandemic.

@@

\hypertarget{socioeconomic-deprivation-and-recovery-of-urban-mobility}{%
\subsection{Socioeconomic deprivation and recovery of urban
mobility}\label{socioeconomic-deprivation-and-recovery-of-urban-mobility}}

In this section we explore further the role of socioeconomic deprivation
in the evolution of the levels of urban mobility. For a given point in
time (i.e.~a month), we start by considering the relationship between
the number of movements relative to the pre-pandemic baseline period and
the average relative deprivation index, at the administrative unit
level. We assume that this relationship is linear and we use a linear
regression to estimate the slope and intercept characterising the line
of best fit. This is shown for April 2020 and March 2022 in the
right-hand side panels of Figure 3. After obtaining the slope and
intercept for every month, we are able to plot the evolution of these
parameters for both the observed and forecasted data, as displayed on
the left-hand-side panels of the same Figure.

\begin{figure}

@@

\end{figure}

We find patterns in the evolution of the estimated parameters that
characterise the relationship between the levels of urban mobility and
RDI. In Argentina, Colombia and Mexico, we observe that the slope of
this relationship evolves to become smaller over time. The tendency is
not apparent in Chile, where the slope of the relationship remains
approximately the same despite the temporary fluctuations. The slope
captures the extent of differences in the level of urban mobility across
administrative units with varying levels of socioeconomic deprivation.
It can therefore be regarded as a measure of inequality in mobility
patterns across socioeconomic groups. A slope equal to zero would mean
that all administrative units display the same intensity of movement
regardless of their socioeconomic deprivation levels. Given the patterns
observed in Argentina, Colombia and Mexico, we find that at the
beginning of the pandemic there were notable inequalities between
socioeconomic groups in terms of the levels of urban mobility. While it
has taken more than two years for Argentina and Mexico to close the gap
(their slope is close to zero from spring 2022), inequalities persist in
Chile and Colombia as of March 2023.

The intercept of the relationship displays stronger patterns, which are
consistent across the four countries. The intercept estimates the urban
mobility levels that would be observed in administrative units where
relative deprivation is zero. The intercept was below the baseline level
at the early stages of the pandemic. As observed in Figure 3, while
there are some differences between countries in the evolution of the
intercept, the general tendency is for the intercept to increase. While
Argentina and Mexico reach values that are closer to the baseline
towards the end of the forecast period, the intercept for Chile and
Colombia remains lower. Therefore, if there were areas with no
socioeconomic deprivation, we would have seen a recovery in the levels
of mobility, especially in Argentina and Mexico

\hypertarget{sec-discussion}{%
\section{Discussion}\label{sec-discussion}}

Using location data from Meta-Facebook users, our study aimed to examine
the evolution of patterns of mobility across socioeconomic groups in
functional urban areas from Argentina, Chile, Colombia and Mexico from
April 2020 to March 2023, following the COVID-19 pandemic. We found a
systematic drop in the number of population movements in April 2020,
with the largest reductions observed in the most affluent administrative
units within functional urban areas from Argentina, Chile and Colombia.
While mobility rebounded closer to pre-pandemic levels approximately two
years later, when COVID-19 restrictions eased, the number of movements
remained below pre-pandemic in Chile. Furthermore, we found that at the
beginning of the pandemic there were inequalities between socioeconomic
groups in terms of the levels of urban mobility. While it has taken more
than two years for Argentina and Mexico to gradually close the gap,
inequalities persist in Chile and Colombia as of March 2023 according to
our estimated data.

We focused the analysis on short-distance movements in urban areas,
especifically those covering 70 km or less. These journeys are typically
considered to represent local and routine mobility (Owen and Green
1992). However, due to the characteristics of the Meta-Facebook movement
data, we are unable to distinguish the purpose of these short-distance
movements. Hence, some of our data could be capturing journeys that
involve a permanent change of place of residence. Our work therefore
motivates the need to answer questions regarding the validity of digital
footprint data for the analysis of human mobility. Further research
should focus on inferring more specific information about the nature of
the journeys, for example following the approach in X and Y (cite
JOana), and quantifying the extent to which the digital footprint data
mirrors the true mobility patterns.

Conducting research on urban mobility using digital footprint data is
not straightforward, due to the challenges in accessing and working with
unstructured data sets which are often subject to biases and statistical
representation issues. These biases often arise from inequalities in
access and usage of digital technologies across demographic groups
(cite). Despite these challenges, the data and analysis that we used for
this work provide evidence for non-trivial patterns that are consistent
across four countries in Latin America and with other parts of the
world. Our findings highlight the dynamic interplay between
socioeconomic status and urban mobility, and shall be used to motivate
and inform the public debate on the deep societal consecuences of how
disparities in urban mobility may be affecting the wider socioeconomic
landscape in Latin American countries.

In conclusion, we argue that this work goes beyond the analysis of
specific patterns by demonstrating the potential of digital footprint
data for policy-relevant research on human mobility at an unprecedented
level of spatiotemporal granularity. While we have seen a rise in
initiatives to improve data services and methodological frameworks to
facilitate the use of digital footprint data for social good, progress
is still limited, especially in some parts of the world including Latin
America. It is in the hands of governments and public organisations to
prioritise maximiasing the societal benefits that digital footprint data
has to offer. This includes engaging in activities such as building
strategic partnerships with commercial data-holders and academic
institutions to establish a unified framework for the use of digital
footprint data in policy and research. In particular, we call for the
creation of resources like those developed by the European Commission
Joint Research Centre (2022) and the UNSD (2019), which identify sources
of non-traditional data and set methodological protocols for
incorporating mobile phone data into official mobility statistics. While
current resources tend to have a global reach, we advocate for more
tailored local initiatives that acknowledge disparities in regional data
availability and adoption of digital technologies.

\hypertarget{sec-methods}{%
\section{Methods}\label{sec-methods}}

\hypertarget{meta-facebook-movement-data}{%
\subsection{Meta-Facebook movement
data}\label{meta-facebook-movement-data}}

>>>>

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\usepackage{lipsum} \usepackage{libertine}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Socioeconomic deprivation and urban mobility in Latin America after COVID-19},
  pdfauthor={Carmen Cabrera-Arnau; Francisco Rowe; Miguel González-Leonardo; Andrea Nasuto; Ruth Neville},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{\textbf{Socioeconomic deprivation and urban mobility in Latin
America after COVID-19}}

@@

\date{}
\begin{document}
\maketitle
\begin{abstract}
The movement of people between locations is crucial for sustainable and
inclusive cities, facilitating the exchange of knowledge and resources.
Due to restrictions, lockdowns and fear of crowded places, the COVID-19
pandemic significantly reduced the number and extent of people's
movements. Existing research suggests that the impact of the pandemic on
mobility was unequal across locations and socioeconomic groups. However,
evidence and analysis of long-term changes in mobility is scarce,
especially in less developed countries. In this study, we use location
data from Meta-Facebook users to analyse patterns of urban mobility in X
Functional Urban Areas from South American countries. Findings reveal a
general decrease in mobility during the pandemic, with gradual recovery
trends up to April 2023. However, socioeconomic disparities are present
through the period of analysis, with deprived areas experiencing smaller
initial losses in mobility. These disparities generally diminish over
time, although they persist in all the countries included in the
analysis. Our research highlights the importance of timely mobility data
with high spatiotemporal resolution to understand the long-term effects
of the pandemic and to inform equitable policy responses that address
societal challenges in urban areas.
\end{abstract}

\section{Introduction}\label{sec-intro}

@@

A key barrier to monitor changes in geographic mobility patterns in less
developed countries during and post the COVID-19 pandemic has been the
lack of suitable data. Traditionally census and survey data have been
employed to analyse human mobility patterns in these countries. Yet,
these data streams are not frequently updated and suffer from slow
releases, with census data for example being collected over intervals of
ten years in most countries. Traditional data streams thus lack the
temporal granularity to analyse population movements over short-time
periods and to offer an up-to-date representation of the urban mobility
system. Data resulting from social interactions on digital platforms
have emerged as an unique source of information to deliver this
representation and capture human population movement in less developed
countries at scale. Particularly location data from mobile phone
applications have become a prominent source to sense patterns of human
mobility at higher geographical and temporal resolution in real time.

Drawing on a dataset of 213 million observations from Meta-Facebook
users' mobile location data, we aim to assess socioeconomic differences
in the extent and persistence of decline in urban mobility in Argentina,
Chile, Colombia and Mexico during and after the COVID-19 pandemic from
March 2020 to March 2023. We use Meta-Facebook data to measure
origin-destination flows from March 2020 to May 2022, and Meta Prophet
time-series forecasting machine learning algorithm to predict
origin-destination flows from June 2022 to March 2023. We use Functional
Urban Areas (FUAs) boundaries from the Global Human Settlement Layer
(GHSL), developed by the European Comission's Joint Research Centre
(Schiavina M. 2019) to define urban areas; and the Global Gridded
Relative Deprivation Index (GRDI) developed by NASA's Socioeconomic Data
and Applications Centre (Columbia University 2022) from sociodemographic
and satellite data inputs. Building on existing evidence (e.g. (Rowe et
al. 2022; Wang et al. 2022)), we hypothesised that (1) urban mobility
has recovered returning to pre-pandemic levels of movement as
nonpharmaceuthical restrictions were lifted; and (2) socioeconomic
differences in urban mobility have endured the pandemic reflecting deep
societal inequalities as knowledge-intensive businesses adopt hybrid
working.

\begin{itemize}
\tightlist
\item
  Context and importance of human mobility: public health , climate
  change\\
\item
  The pandemic resulted in major changes in human mobility - unequal
  impacts across socioeconomic groups\\
\item
  Gap 1: Little work on understanding the long-term changes in mobility
  - recovery\\
\item
  Gap 2: little work on less developed countries - where inequalities
  are more pronounced\\
\item
  Aim - focus on Latin American countries as case study.
\item
  Aim - focus on urban areas, as the future of humanity is the future of
  cities and Latin Ameriaca is highly urbanised
\end{itemize}

\section{Results}\label{sec-results}

The evolution of the levels of movement is measured with respect to a
baseline period prior to the pandemic as described in
Section~\ref{sec-methods}. For the purposes of the analysis, we
aggregate the raw movement data temporally into months and spatially
into administrative units at various levels according to GADM, the
Database of Global Administrative Areas (GADM 2024). The analysis
focuses on administrative units that are within the boundaries of
functional urban areas as specified by the Global Human Settlement Layer
(GHSL). For each administrative unit, we compute the Relative
Deprivation Index (RDI) based on data from NASA's Socioeconomic Data and
Applications Centre (SEDAC). Figure 1 displays the administrative units
included in the study, coloured according to their average RDI.
Predictions about the evolution of the levels of movement are made using
the Prophet forecasting procedure. Further details are provided in
Section~\ref{sec-methods}.

\begin{figure}[H]

@@

\end{figure}%

\subsection{The heterogeneous impact of COVID-19 on urban
mobility}\label{the-heterogeneous-impact-of-covid-19-on-urban-mobility}

We analyse the evolution of the relative intensity of urban mobility
with respect to a baseline period prior to the pandemic. Specifically,
we focus the analysis on short-distance movements in urban areas to
represent local and routine mobility (Owen and Green 1992), so only
movements covering a distance of at most 70 km are considered. For a
movement to be classified as urban it needs to start or end within a
functional urban area from Argentina, Chile, Colombia and Mexico. The
observed data is available for a two-year period starting in April 2020,
just after the first wave of COVID-19 pandemic cases, and ending in
March 2022. After 2022 no observations are available, however, we
generate a 12-month forecast up to March 2023 in order to gain a better
understanding of the recovery trends.

Figure 2 displays the patterns of recovery for the mobility levels in
the administrative units belonging to functional urban areas in the
countries of interest. The three lines in each panel represent the mean
levels of mobility for administrative units grouped into one of three
terciles, according to their average RDI.

\begin{figure}[H]

@@

\end{figure}%

Generally, there was a drop in the levels of mobility with respect to
the baseline period in all four countries. This drop was especially
large for Argentina, Chile and Colombia, with Mexico displaying a
smaller decrease in the relative number of movements with respect to the
baseline. Following the initial drop in mobility, all four countries
evolve towards the recovery of baseline levels, with a generally
increasing trend. There are, however, fluctuations from the general
trend which manifest differently for each country. These fluctuations
mirror each other in the case of Argentina and Colombia, where urban
mobility sharply bounces back closer to pre-pandemic levels around July
of 2020. Chile and Mexico display more progressive patterns of recovery,
although Chile never reaches baseline levels. These fluctuations are
unique to each country and can be attributed to local factors such as
the effects of seasonality or the different stringency measures imposed
by the national governments during the pandemic.

@@

\subsection{The most deprived areas experienced the smallest drop in
mobility relative to pre-pandemic
levels}\label{the-most-deprived-areas-experienced-the-smallest-drop-in-mobility-relative-to-pre-pandemic-levels}

In this section we explore further the role of socioeconomic deprivation
in the evolution of the levels of urban mobility. For a given point in
time (i.e.~a month), we start by considering the relationship between
the number of movements relative to the pre-pandemic baseline period and
the average RDI, at the administrative unit level. We assume that this
relationship is linear and we use a linear regression to estimate the
slope and intercept characterising the line of best fit. This is shown
for April 2020 and March 2022 in the right-hand side panels of Figure 3.
After obtaining the slope and intercept for every month, we are able to
plot the evolution of these parameters for both the observed and
forecasted data, as displayed on the left-hand-side panels of the same
Figure.

\begin{figure}[H]

@@

\end{figure}%

We find patterns in the evolution of the estimated parameters that
characterise the relationship between the levels of urban mobility and
RDI. In Argentina, Colombia and Mexico, we observe that the slope of
this relationship evolves to become smaller over time. The tendency is
not apparent in Chile, where the slope of the relationship remains
approximately the same over time despite the temporary fluctuations. The
slope captures the extent of differences in the level of urban mobility
across administrative units with varying levels of socioeconomic
deprivation. It can therefore be regarded as a measure of inequality in
mobility patterns across socioeconomic groups. A slope equal to zero
would mean that all administrative units display the same intensity of
movement regardless of their average RDI. Given the patterns observed in
Argentina, Colombia and Mexico, we find that at the beginning of the
pandemic there were notable inequalities between socioeconomic groups in
terms of the levels of urban mobility. While it has taken more than two
years for Argentina and Mexico to close the gap (their slope is close to
zero from spring 2022), inequalities persist in Chile and Colombia as of
March 2023.

The intercept of the relationship displays stronger patterns, which are
consistent across the four countries. The intercept estimates the urban
mobility levels that would be observed in administrative units where the
RDI is zero. The intercept was below the baseline level at the early
stages of the pandemic. As observed in Figure 3, while there are some
differences between countries in the evolution of the intercept, the
general tendency is for the intercept to increase. While Argentina and
Mexico reach values that are closer to the baseline towards the end of
the forecast period, the intercept for Chile and Colombia remains lower.
Therefore, if there were areas with no socioeconomic deprivation, we
would have seen a recovery in the levels of mobility, especially in
Argentina and Mexico

\section{Discussion}\label{sec-discussion}

Using location data from Meta-Facebook users, our study aimed to examine
the evolution of patterns of mobility across socioeconomic groups in
functional urban areas from Argentina, Chile, Colombia and Mexico from
April 2020 to March 2023, following the COVID-19 pandemic. We found a
systematic drop in the number of population movements in April 2020,
with the largest reductions observed in the most affluent administrative
units within functional urban areas (FUAs) from Argentina, Chile and
Colombia. While mobility rebounded closer to pre-pandemic levels
approximately two years later, when COVID-19 restrictions eased, the
number of movements remained below pre-pandemic in Chile. Furthermore,
we found that at the beginning of the pandemic there were inequalities
between socioeconomic groups in terms of the levels of urban mobility.
While it has taken more than two years for Argentina and Mexico to
gradually reduce gap, inequalities persist as of March 2023, especially
in Chile and Colombia according to our estimated data.

We focused the analysis on short-distance movements in urban areas,
especifically those covering 70 km or less. These journeys are typically
considered to represent local and routine mobility (Owen and Green 1992)
. However, due to the characteristics of the Meta-Facebook movement
data, we are unable to distinguish the purpose of these short-distance
movements. Hence, some of our data could be capturing journeys that
involve a permanent change of place of residence. Our work therefore
motivates the need to answer questions regarding the validity of digital
footprint data for the analysis of human mobility. Further research
should focus on inferring more specific information about the nature of
the journeys, following similar approaches to those proposed by
Cabrera-Arnau et al. (2023), and quantifying the extent to which the
digital footprint data mirrors the true mobility patterns.

Conducting research on urban mobility using digital footprint data is
not straightforward, due to the challenges in accessing and working with
unstructured data sets which are often subject to biases and statistical
representation issues. These biases often arise from inequalities in
access and usage of digital technologies across demographic groups (Rowe
2023). Despite these challenges, the data and analysis that we used for
this work provide evidence for non-trivial patterns that are consistent
across four countries in Latin America and with other parts of the
world. Our findings highlight the dynamic interplay between
socioeconomic status and urban mobility, and shall be used to motivate
and inform the public debate regarding the deep societal consequences of
urban mobility disparities on the wider socioeconomic landscape of Latin
American countries.

In conclusion, we argue that this work goes beyond the analysis of
specific patterns by demonstrating the potential of digital footprint
data for policy-relevant research on human mobility at an unprecedented
level of spatiotemporal granularity. While we have seen a rise in
initiatives to improve data services and methodological frameworks to
facilitate the use of digital footprint data for social good, progress
is still limited, especially in some parts of the world including Latin
America. It is in the hands of governments and public organisations to
prioritise the maximisation the societal benefits that digital footprint
data has to offer. This includes engaging in activities such as building
strategic partnerships with commercial data-holders and academic
institutions to establish a unified framework for the use of digital
footprint data in policy and research. In particular, we call for the
creation of resources like those developed by the European Commission
Joint Research Centre (Commission et al. 2022) and the UN Statistics
Division (Division 2019), which identify sources of non-traditional data
and set methodological protocols for incorporating mobile phone data
into official mobility statistics. While current resources tend to have
a global reach, we advocate for more tailored local initiatives that
acknowledge disparities in regional data availability and adoption of
digital technologies.

\section{Methods}\label{sec-methods}

\subsection{Meta-Facebook movement
data}\label{meta-facebook-movement-data}

To capture population movements, we used anonymised aggregate mobile
phone location data from Meta users for Argentina, Colombia, Chile and
Mexico, covering a 24-month period from April 2020 to March 2022. We
used the dataset Facebook Movements created by Meta and accessed through
their Data for Good Initiative (\url{https://dataforgood.facebook.com}).
The data are built from Facebook app users who have the location
services setting turned on on their smartphone. Prior to releasing the
datasets, Meta ensures privacy and anonymity by removing personal
information and applying privacy-preserving techniques (Maas et al.
2019). Small-count dropping is one of these techniques. A data entry is
removed if the population or movement count for an area is lower than
10. The removal of small counts may mean that population counts in small
sparsely populated areas are not captured. A second technique consists
in adding a small undisclosed amount of random noise to ensure that it
is not possible to ascertain precise, true counts for sparsely populated
locations. Third, spatial smoothing using inverse distance-weighted
averaging is also applied is applied to produce a smooth population
count surface. The Facebook Movements dataset offers information on the
total number of Facebook users moving between and within spatial units
in the form of origin-destination matrices. The data is temporally
aggregated into three daily 8-hour time windows (i.e.~00:00-08:00,
08:00-16:00 and 16:00-00:00). The dataset includes a baseline capturing
the number of movements before COVID-19 based on a 45-day period ending
on March 10th 2020. The baseline is computed using an average for the
same time of the day and day of the week in the period preceding March
10th. For instance, the baseline for Monday 00:00-08:00 time window is
obtained by averaging over data collected on Mondays from 00:00 to 8:00
for the 45-day period. Details about the baseline can be found in (Maas
et al. 2019). The Bing Maps Tile System developed by Microsoft
(Microsoft) is used a spatial framework to organise the data. The Tile
System is a geospatial indexing system that partitions the world into
tile cells in a hierarchical way, comprising 23 different levels of
detail (Microsoft). At the lowest level of detail (Level 1), the world
is divided into four tiles with a coarse spatial resolution. At each
successive level, the resolution increases by a factor of two. The data
that we used are spatially aggregated into Bing tile levels 13. That is
about 4.9 x 4.9km at the Equator (Maas et al. 2019).

\subsection{Spatiotemporal data
aggregation}\label{spatiotemporal-data-aggregation}
